Analysis of competitive surfing tournaments with generalized Bradley-Terry

### 2021-08-23 inst/surfing_jsa.tex
work on Tim's suggestions

The statistical method described here can test many different
suggestions drawn from surfing lore.  For example, one frequent
suggestion~\cite{burgess2020,ho2021} is that Brazilian surfers such as
Ferreira and Toledo tend to have surfing skillsets that favour point
and beach break beaches {\bf Tim I need some help describing and
  defining these here}, those types of beach being readily accessible
to these surfers' native Brazil.  We now assess the null that
Brazilians perform equally well on all types of beaches.  To that end,
we create a reified surfer, ``Brazilian beachtype", whose strength
helps all Brazilian surfers when surfing on their (putative) preferred
beach type, an analogue of the ``home ground monster" who helps the
home team as per equation~\ref{home_ground_advantage}.  Here we modify
the likelihood function only when a single Brazilian is in the water,
to avoid possible interaction with the compatriot issue discussed in
section~\ref{compatriot}.  The null would be that this reified entity
has zero strength.  We fail to reject this null (differential support
= 0.035; $p=0.79$) and conclude that there is no evidence to suggest
that Brazilians' performce is better on their preferred beach types
than one would expect by chance.  We may go further using the method
of support: not only is the maximum likelihood estimate for Brazilian
beachtype very small (about 0.002), but in addition we may reject the
the proposal that the strength of this entity exceeds about 0.025
(Figure~\ref{brazilian_beachtype_support}).

@@

Consider, as an example, the case of Smith, Buchan, Freestone who were
simultaneously in the water at Bells Beach, Australia.  Now, Buchan
and Freestone are Australian, and Jordy Smith is South African.  It is
reasonable to ask whether Buchan and Freestone, who are in direct
competition with one another for a place in the Australian Olympic
surfing squad, would be more motivated to beating each other than to
come ahead of Smith.  This would be a rational response, but arguably
noncompetitive.  How would such a strategy manifest itself in
scorelines?

Reified Bradley-Terry technique furnishes an intuitive statistical
test for mechanisms that control for competitors' strengths.  Noting
that the compatriots effectively place themselves at a {\em
  disadvantage} when compared with the odd man out, we introduce a
compatriot monster whose strength $p_C$ helps the odd man out.
Writing $p_1, p_2$ for the strengths of the compatriots and $q$ for
the odd man out, we construct likelihood functions for the six
possible orderings when two compatriots and an odd man out are in the
water at the same time in Table~\ref{likelihoodcompatriot} (if we have
either three different nationalities or three identical nationalities,
we revert to the standard Plackett-Luce likelihood).  Note that all
six likelihood functions in table~\ref{likelihoodcompatriot} are
informative about $p_1,p_2,q$ in addition to being informative about
the object of inference, $p_C$.

>>>>

The statistical method described here can test many different
hypotheses drawn from surfing lore.  For example, one frequent
suggestion~\cite{burgess2020,ho2021} is that Brazilian surfers such as
Ferreira and Toledo tend to have skillsets that favour point and beach
break beaches, those types of beach being readily accessible to these
surfers' native Brazil; see~\cite{scarfe2003} for a scientific
overview of surf break characteristics, and~\cite{butt2004} for a
surfer's perspective.

We now assess the null that
Brazilians perform equally well on all types of beaches.  To that end,
we create a reified surfer, ``Brazilian beachtype", whose strength
helps all Brazilian surfers when surfing on their (putative) preferred
beach type, an analogue of the ``home ground monster" who helps the
home team as per equation~\ref{home_ground_advantage}.  Here we modify
the likelihood function only when a single Brazilian is in the water,
to avoid possible interaction with the compatriot issue discussed in
section~\ref{compatriot}.  The null would be that this reified entity
has zero strength.  We fail to reject this null (differential support
= 0.035; $p=0.79$) and conclude that there is no evidence to suggest
that Brazilians' performce is better on their preferred beach types
than one would expect by chance.  We may go further using the method
of support: not only is the maximum likelihood estimate for Brazilian
beachtype very small (about 0.002), but in addition we may reject the
the proposal that the strength of this entity exceeds about 0.025
(Figure~\ref{brazilian_beachtype_support}).

@@

Consider, as an example, the case of Smith, Buchan, Freestone who were
simultaneously in the water at Bells Beach, Australia.  Now, Buchan
and Freestone are Australian, and Jordy Smith is South African.  It is
reasonable to ask whether Buchan and Freestone, who are in direct
competition with one another for a place in the Australian Olympic
surfing squad, would be more motivated to beating each other than to
come ahead of Smith.  This would be a rational response, but arguably
noncompetitive.  How would such a strategy manifest itself in
scorelines?  We would suggest that Smith is able to
modify his strategy to benefit from the mutual rivalry between
Buchan and Freestone; and indeed we see Smith winning that particular
heat.

Reified Bradley-Terry technique furnishes an intuitive statistical
test for such mechanisms, that automatically controls for competitors'
strengths.  Noting
that the compatriots effectively place themselves at a {\em
  disadvantage} when compared with the odd man out, we introduce a
compatriot monster whose strength $p_C$ helps the odd man out.
Writing $p_1, p_2$ for the strengths of the compatriots and $q$ for
the odd man out, we construct likelihood functions for the six
possible orderings when two compatriots and an odd man out are in the
water at the same time in Table~\ref{likelihoodcompatriot} (if we have
either three different nationalities or three identical nationalities,
we revert to the standard Plackett-Luce likelihood).  Note that all
six likelihood functions in table~\ref{likelihoodcompatriot} are
informative about $p_1,p_2,q$ in addition to being informative about
the object of inference, $p_C$.
